---
title: "Exploratory Analysis"
output: html_document
---

## Dog Adoptions!

###Ricky Le, Nalin Udomsak, Ky Truong, Chantria Im

### February 20th, 2023

### Abstract
Our main question is what are the features of dogs in shelters across the United States? To address the question, we will be using data collected across all shelters on a single day in 2019. This is important because it can provide an insight into key characteristics of dogs in shelters and others can use this data to help more dogs get adopted. 

### Keywords
Dog Adoptions, Maps, Pie Chart, Bar Chart

### Introduction  

Briefly introduce your project.  Include 3-5 research questions. What motivates the questions? Why are they important? (at least 200 words)

### Related Work  
According to the [American Society for the Prevention of Cruelty to Animals](https://www.aspca.org/helping-people-pets/shelter-intake-and-surrender/pet-statistics), approximately 6.5 million animals enter shelters each year and only 3.2 million are adopted.  Research shows that trends towards dog adoption and shelters have changed since the late 1960s largely due to a cultural and social shift in how we relate to dogs which have led to a sustaining decline in unwanted animals entering shelters and being euthanized. This includes the role of sterlization efforts, increased responsible pet ownership behaviors like their level of care and control [(Rowan & Kartal, 2019)](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5981279/). For animals in shelters, data collected from PetPoint shows that the adoption rate has also increased over steadily over the past decade. An article titled “Characterizing unsuccessful animal adoptions: age and breed predict the likelihood of return, reasons for return and post-return outcomes” by Powell, found that dogs were largely returned to the shelter due to behavoral problems. This includes barking, hyperactivity, and aggression towards humans and animals. For key characteristics among dogs that were returned, Powell found that young adult dogs and pit bull-type breeds made up the largest portion [(Powell et al., 2021)](https://www.nature.com/articles/s41598-021-87649-2). Lastly, an article titled “Increasing adoption rates at animal shelter: a two-phase approach to predict length of stay and optimal shelter allocation” by Bradley et al., found that the age, multicolor and size of the dogs were important factors to how long it would take for the dog to be adopted [(Bradley et al., 2021)](https://bmcvetres.biomedcentral.com/articles/10.1186/s12917-020-02728-2).

*References*

Rowan, A., & Kartal, T. Dog Population & Dog Sheltering Trends in the United States of America. *Animals : an open access journal from MDPI,* 8(5), 68 (2018). 

Powell, L., Reinhard, C., Satriale, D. et al. Characterizing unsuccessful animal adoptions: age and breed predict the likelihood of return, reasons for return and post-return outcomes. *Sci Rep 11,* 8018 (2021).

Bradley, J., Rajendran, S. Increasing adoption rates at animal shelters: a two-phase approach to predict length of stay and optimal shelter allocation. *BMC Vet Res 17,* 70 (2021). 

### The Dataset

Where did you find the data? Please include a link to the data source  
Who collected the data?  
How was the data collected or generated?  
Why was the data collected?  
How many observations (rows) are in your data?  
How many features (columns) are in the data?  
What, if any, ethical questions or questions of power do you need to consider when working with this data?  
What are possible limitations or problems with this data?   (at least 200 words)

### Implications

Assuming you answer your research questions, briefly describe the expected or possible implications for technologists, designers, and policymakers. (at least 150 words)

### Limitations & Challenges
What challenges or limitations might you need to address with your project idea more broadly? Briefly discuss. (at least 150 words)

### Summary Information

Write a summary paragraph of findings that includes the 5 values calculated from your summary information R script

### Table

Include a table of aggregate information

Describe why you included the table and what information it reveals

### Chart 1

Include a chart

Describe why you chose this chart and what information it reveals

Here's an example of how to run an R script inside an RMarkdown file:

```{r Chart3.1_nalin.R , fig.width=12}
bar_chart
```
This bar chart displays the distribution of dogs in Shelters by their primary color on a given day in September, 2019. I chose to display this data through a bar chart because it uses categorical data with many categories. This data visualization means that each bar represents the count of that category. The bar chart shows that dogs with black fur were the most popular with over 7000 counts in the dataset. This was followed by dogs with white and brown fur. Whilst dogs with harlequin fur were the least popular with 73 counts. 

### Chart 2

Include a chart

Describe why you chose this chart and what information it reveals

### Chart 3

Include a chart

Describe why you chose this chart and what information it reveals

```{r chart1_chantria.R , fig.width=12}
map
```
We chose a map to represent the distribution of adoptable dogs across the United States because it provides a visualization of which states have a high and low concentration of dogs sitting in shelters. Based on the visualization, it appears that many states within the Rocky Mountain and Midwest regions have a high volume of dogs sitting in shelters compared to urban areas, such as Washington, California and New York. We can conclude that more urbanized and populated areas are likely to have better resources and have a larger pool of potential adopters compared to rural areas have more supply than demand. It is important to have relocation programs in place to ensure that dogs are distributed to areas where they are more likely to find homes.
